An Efficient Minibatch Acceptance Test for MCMC

### 2016-07-05 nips_2016.tex
updates to first section

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$

The acceptance test guarantees detailed balance, so that
\begin{equation}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta_t,\theta') = 
\min\left\{ 1, \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
\theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)} \right\},
\end{equation}
where $p(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points. Moreover, it is difficult to
design tests using substantially fewer than $N$ points that also
satisfy detailed balance. To reconcile these competing
objectives,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta_t,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta_t,\theta')$ in
the equivalent form $\Lambda(\theta_t,\theta') > \psi(u,\theta_t,\theta')$ where
\begin{equation}
\Lambda(\theta_t,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta_t)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta_t,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta_t)p(\theta_t)}{q(\theta_t|\theta')p(\theta')}\right)
\end{equation}

By the central limit theorem, we expect $\Lambda(\theta_t,\theta')$ to be approximately
Normal. To reduce computational effort, an unbiased estimate of $\Lambda(\theta_t,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta_t,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta_t)}\right)  
\end{equation}

The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta_t,\theta') > \psi(u,\theta_t,\theta')$ by
establishing that $\Lambda^*(\theta_t,\theta')$ is substantially larger
than $\psi(u,\theta_t,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instance consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(N)$, where $N$ is the number of data points.

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta_t,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta_t)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta_t)^2\right) =
  (\theta'-\theta_t)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta_t-\frac{\theta'-\theta_t}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta_t)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta_t)$
and $\psi(u,\theta',\theta_t)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\theta'-\theta_t,1/N)$, so we take $q(\theta'\mid\theta_t) =
\phi((\theta'-\theta_t)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta_t)=q(\theta_t,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta_t)$ reduces to $1/N\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta_t, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

\begin{proof} Assume $(\theta' - \theta_t) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and
  $\theta_t -0.5 \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events
  occur with probability $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and $p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively,
  and guarantee that $\Lambda^*(\theta',\theta_t)$ is negative, which guarantees that we can find
  a $u$ so that $\psi(u,\theta',\theta_t)$ equals the expected value of $\Lambda^*(\theta',\theta_t)$.
  Specifically, choose $u_0$
  to satisfy
  \begin{equation}
    \log u_0 = N(\Lambda^*(\theta',\theta_t)-\psi(1,\theta',\theta_t))
  \end{equation}
  which evaluates to 
  \begin{equation}
    \log u_0 = -N(\theta'-\theta_t)\left(\theta_t-0.5+\frac{\theta'-\theta_t}{2}\right) 
  \end{equation}
  and rewrite the test now as
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta_t+\frac{\theta'-\theta_t}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right)
    \hspace{0.1in} \not\approx \hspace{0.1in} 0
  \end{equation}
  where $\not\approx$ means ``significantly different from'' under the distribution over samples of $x_i$. 
  The above choice of $u_0$ ensures that the terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the
  test will never correctly succeed.

@@

 The choices above for $\theta_t$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
  \end{equation}
  the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. With $u$ in this range, we rewrite the test as:
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
  \end{equation}
  so that the LHS has expected value zero.  Given our choice of intervals for the variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.
  The standard deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$. And
  so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

  The samples $\theta_t$, $(\theta'-\theta_t)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios as $\Delta$ and $\Delta^*$.
First, $\Delta$ is defined as:
\begin{equation}\label{eq:delta0}
    \Delta(x,\theta_t,\theta') =
\log \frac{p(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta_t \mid
  \theta')}{p(\theta_t)\prod_{i=1}^N p(x_i \mid \theta_t)q(\theta' \mid\theta_t)}
  \end{equation}
and we can separate out terms dependent and independent of the data $x$ as:
\begin{equation}\label{eq:delta1}
\Delta(x,\theta_t,\theta') = N\psi(1,\theta_t,\theta') + \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta_t)}
\end{equation}
\begin{equation}\label{eq:delta2}
    \Delta^*(x,\theta_t,\theta') =
N\psi(1,\theta_t,\theta') + \frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta_t)}
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a minibatch of size $b$
respectively. The scaling term $N$ produces true log likelihood ratios whereas $\Lambda-\psi$
is a scaled log ratio. $\Delta^*$ is an unbiased estimator of $\Delta$.

\subsection{A Full Data Acceptance Test}\label{ssec:deltas_full}

The key to our test is to smooth the acceptance function.  We consider tests other than the Metropolis
test that satisfy the conditions needed for accurate posterior estimation. This condition is
the detailed balance condition defined earlier. Expressed in terms of $\Delta$, the
classical Metropolis test is $f(\Delta) = {\rm min}(\exp(\Delta),1)$. More generally we have:

\begin{lemma}\label{lem:detailed_balance}
An acceptance function $g$ such that $g(\Delta) = \exp(\Delta) g(-\Delta )$ satisfies detailed
balance. That is, $f(\theta_t)p(\theta' \mid \theta_t) = f(\theta')p(\theta_t \mid \theta')$, where
$p(\theta_y \mid \theta_x)$ is the probability of jumping from $\theta_x$ to $\theta_y$ in the
chain.
\end{lemma}

\begin{proof}
First note that the transition probability $p(\theta' \mid \theta_t)$ is the probability of proposing
$\theta'$ and then accepting it, so $p(\theta' \mid \theta_t) = q(\theta' \mid \theta_t)g(\Delta).$
Similarly, $p(\theta_t \mid \theta') = q(\theta_t \mid \theta')g(-\Delta).$ Notice that the
probability of accepting a transition from $\theta'$ to $\theta_t$ is $g(-\Delta)$ because this
inverts the fraction inside the logarithm term of $\Delta$.  By assumption, we can expand $g(\Delta)
= \exp(\Delta)g(-\Delta)$ in $p(\theta' \mid \theta_t)$. Doing this, and combining the result with
the definition of $p(\theta_t \mid \theta')$, we get
\begin{equation}\label{eq:combined}
g(-\Delta) = \frac{p(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t)\exp(\Delta)} = \frac{p(\theta_t \mid \theta')}{q(\theta_t \mid \theta')}.
\end{equation}
Rearranging terms and expanding $\exp(\Delta)$, we have
\begin{equation}\label{eq:rearrange}
\frac{p(\theta' \mid \theta_t) f(\theta_t) q(\theta' \mid \theta_t)}{q(\theta' \mid \theta_t) f(\theta') q(\theta_t \mid \theta')} = \frac{p(\theta_t \mid \theta')}{ q(\theta_t \mid \theta')}.
\end{equation}
Cancellations result in $f(\theta') p(\theta_t \mid \theta') = f(\theta_t) p(\theta' \mid
\theta_t)$. Thus, detailed balance is satisfied.
\end{proof}

As a sanity check, the standard Metropolis-Hastings acceptance function $g(\Delta) = \min\{1,
e^\Delta \} = \min\left\{1, \frac{f(\theta')q(\theta_t \mid \theta')}{f(\theta_t)q(\theta' \mid
  \theta_t)}\right\}$ satisfies the condition $g(\Delta) = \exp(\Delta)g(-\Delta)$. In fact it is
the optimal acceptance function in terms of acceptance rate, since it accepts with probability
1 for $\Delta > 0$.

\subsubsection{Logistic Acceptance Function}\label{ssec:deltas_full}
For our new MH test we use the logistic function: $g(\Delta) =
(1+\exp(-\Delta))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  The logistic function is nice because we can easily sample from
it using the following procedure. At any iteration with a current parameter $\theta_t$ and a
candidate sample $\theta'$, we can compute $\Delta$. Let $u$ be a uniform random variable $u \sim
{\rm Unif}[0,1]$. We accept $\theta'$ if $g(\Delta) > u$, and reject otherwise. This process is
equivalent to sampling a random variable $X$ with cumulative distribution function $F_X(x) = g(x)$,
and accepting if $\Delta > X$ and rejecting otherwise. We can define $X = g^{-1}(u)$ so that its CDF
is the logistic function, which follows because for arbitrary $X = x$, we have
\[
F_X(x) = {\rm Pr}(X \le x) = {\rm Pr}(g^{-1}(u) \le x) = {\rm Pr}(u \le g(x)) = \int_{0}^{g(x)} 1 dx = g(x),
\]
as the density of $u$ is one. Thus, the criteria to accept the candidate $\theta'$ is equivalent to
whether $\Delta > X$. Moreover, the density of $X$ (a standard logistic random variable) is
symmetric about zero, so the acceptance criteria can also be expressed as $\Delta + X>0$.

>>>>

\usepackage[utf8]{inputenc} % allow utf-8 input
\usepackage[T1]{fontenc}    % use 8-bit T1 fonts
\usepackage{hyperref}       % hyperlinks
\usepackage{url}            % simple URL typesetting
\usepackage{booktabs}       % professional-quality tables
\usepackage{amsfonts}       % blackboard math symbols
\usepackage{amssymb}       % blackboard math symbols
\usepackage{nicefrac}       % compact symbols for 1/2, etc.
\usepackage{microtype}      % microtypography
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}

@@

In the Metropolis-Hastings method~\cite{gilks1996markov,brooks2011handbook}, a difficult-to-compute
probability distribution $p(\theta)$ is sampled using a Markov chain $\theta_1,\ldots,\theta_n$. The 
sample $\theta_{t+1}$ at time $t+1$ is generated using a candidate $\theta'$ from a (simpler) proposal
distribution $q(\theta'\mid \theta_t)$, filtered by an acceptance test. The acceptance test is usually
a Metropolis test. The Metropolis test has acceptance probability:
\begin{equation}\label{eq:traditional}
\alpha(\theta_t,\theta') = \min\left\{ 1, \frac{p(\theta')q(\theta_t \mid \theta')}{p(\theta_t)q(\theta' \mid \theta_t)}
\right\}
\end{equation}
and with probability $\alpha(\theta_t,\theta')$, we accept $\theta'$
and set $\theta_{t+1} = \theta'$, otherwise set
$\theta_{t+1}=\theta_t$.  The test is often implemented with an
auxiliary random variable $u \sim \mathcal{U}(0,1)$ with a comparison $u < \alpha(\theta_t,\theta')$.
For notational simplicity, from now on we will drop the subscript $t$ for the current sample $\theta_t$ and
denote is simply as $\theta$.

The acceptance test guarantees detailed balance, which means
\begin{equation}\label{detailed_balance1}
  p(\theta)p(\theta,\theta') = p(\theta')p(\theta',\theta)
\end{equation}
where $p(\theta,\theta')$ is the probability of a transition from state $\theta$ to $\theta'$. Here
$p(\theta,\theta') = q(\theta'\mid\theta)\alpha(\theta,\theta')$ so the detailed balance equation becomes
\begin{equation}\label{detailed_balance2}
  p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
\end{equation}
The detailed balance condition, together with ergodicity, guarantee that the
Markov chain has a unique stationary distribution $\pi(\theta) = p(\theta)$.

For Bayesian inference, we would like to sample from a parameter distribution for $\theta$ based on some
observed data $x_1,\ldots,x_N$. i.e. we want to sample from 
$p(\theta \mid x_1, \ldots, x_N)$. The acceptance probability is now:
\begin{equation}
\alpha(\theta,\theta') = 
\min\left\{ 1, \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
\theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)} \right\},
\end{equation}
where $p_0(\theta)$ is a prior, and $p(x_i \mid \theta)$ are the
probabilities of the observations. Computing samples this way requires
the use of all $N$ training data points, but this is very expensive
for large datasets. To address this challenge,~\cite{cutting_mh_2014,icml2014c1_bardenet14} perform
approximate Metropolis-Hasting tests using sequential hypothesis
testing. During each iteration, they start with a small minibatch of
data and test the hypothesis that the sample $\theta'$ based on an
approximate version of the test $u < \alpha(\theta,\theta')$. If the
approximate test cannot make a decision with sufficient confidence,
then the minibatch size is increased and the test repeats. This
process continues until a decision. The bounds depend on either a
asymptotic central limit theorem \cite{cutting_mh_2014} or a
concentration bound \cite{icml2014c1_bardenet14}, the latter requiring
direct bounds on the log likelihood ratio.  The problem with both
methods is the expense of resolving small differences.  In the worst
case, all $N$ data points may be needed, and as we show next this
worst case can occur about $\Omega(N)$ times during the performance of
$N$ tests.

Following ~\cite{icml2014c1_bardenet14}, we write the test $u < \alpha(\theta,\theta')$ in
the equivalent form $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ where
\begin{equation}
\Lambda(\theta,\theta') = \frac{1}{N}\sum_{i=1}^N \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\hspace{0.2in} {\rm and} \hspace{0.2in}
\psi(u,\theta,\theta') = \frac{1}{N}\log\left(u\frac{q(\theta'|\theta)p_0(\theta)}{q(\theta|\theta')p_0(\theta')}\right)
\end{equation}

To reduce computational effort, an unbiased estimate of $\Lambda(\theta,\theta')$
based on a minibatch can be used:

\begin{equation}
\Lambda^*(\theta,\theta') = \frac{1}{b}\sum_{i=1}^b \log\left(\frac{p(x_i|\theta')}{p(x_i|\theta)}\right)  
\end{equation}

The minibatch should contain randomly selected samples, and
by the central limit theorem, we expect $\Lambda^*(\theta,\theta')$ to be approximately
Normal. 
The acceptance test then becomes a statistical test of the hypothesis
that $\Lambda(\theta,\theta') > \psi(u,\theta,\theta')$ by
establishing that $\Lambda^*(\theta,\theta')$ is substantially larger
than $\psi(u,\theta,\theta')$.  In ~\cite{cutting_mh_2014} an
asymptotic central limit argument was used to derive this gap, while
in \cite{icml2014c1_bardenet14} a concentration bound was used. In
both cases, the resulting tests were shown to give useful reductions
in number of samples required over using the full dataset. But no
worst-case bounds were given on average number of samples needed.

We next show that for some simple distributions, the lower
bound of average number of data instances consumed for one iteration of
~\cite{cutting_mh_2014} and ~\cite{icml2014c1_bardenet14} is
$\Omega(N)$, where $N$ is the number of data points.

Consider a simple model where samples $x_i$ for $i=1,\ldots,N$ are
drawn from a Normal distribution $\mathcal{N}(\theta,1)$ where
$\theta=0.5$, and with a uniform prior on $\theta$. The log likelihood ratio is
\begin{equation}
\Lambda^*(\theta,\theta') =  \frac{1}{b}\sum_{i=1}^b \log\frac{p(x_i|\theta')}{p(x_i|\theta)}=
  %%\frac{1}{2}\left( (x_i-\theta')^2 - (x_i - \theta)^2\right) =
  (\theta'-\theta)\left(\frac{1}{b}\sum_{i=1}^b x_i-\theta-\frac{\theta'-\theta}{2}\right)
\end{equation}
which is normally distributed over selection of the Normal samples $x_i$.
Since the $x_i$ have unit variance, their mean has variance $1/b$, and the variance $\sigma^2(\Lambda^*)$
of $\Lambda^*(\theta,\theta')$ is $(\theta'-\theta)^2/b$. In order to pass
a hypothesis test that $\Lambda > \psi$, we will need to establish a
large enough gap (several $\sigma(\Lambda^*)$) between $\Lambda^*(\theta',\theta)$
and $\psi(u,\theta',\theta)$.

We would like to sample from the posterior distribution of $\theta$
which is a normal distribution centered on the sample mean, and with
variance $1/N$.  The target distribution is therefore
$\mathcal{N}(\mu,1/N)$ where $\mu$ is the sample mean.  An efficient
proposal distribution has the same variance as the target distribution
$\mathcal{N}(\theta'-\theta,1/N)$, so we take $q(\theta'\mid\theta) =
\phi((\theta'-\theta)N)$, where $\phi(x)$ is the Normal density
function with zero mean and unit variance. This proposal is symmetric
$q(\theta',\theta)=q(\theta,\theta')$, and we assumed uniform
prior for $\theta$, so $\psi(u,\theta',\theta)$ reduces to $1/N\log
u$.

\begin{lemma}\label{lem:efficienty_cutting}
  For the model above, there
  exists a fixed (independent of $N$) constant $c$ such that with probability $ \geq c$ over the joint
  distribution of $(\theta, \theta', u)$, the
  test from \cite{icml2014c1_bardenet14} consumes all $N$ samples. 
\end{lemma}

\begin{proof} Assume $(\theta' - \theta) \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ and
  $\theta -0.5 \in \pm\frac{1}{\sqrt{N}}[0.5,1]$ with matching sign. These events
  occur with probability $p_0=2*(\Phi(1)-\Phi(0.5)) = 0.2997$ and $p_1=(\Phi(1)-\Phi(0.5))=0.14988$ respectively,
  and guarantee that $\Lambda^*(\theta',\theta)$ is negative, which guarantees that we can find
  a $u$ so that $\psi(u,\theta',\theta)$ equals the expected value of $\Lambda^*(\theta',\theta)$.
  Specifically, choose $u_0$
  to satisfy
  \begin{equation}
    \log u_0 = N(\Lambda^*(\theta',\theta)-\psi(1,\theta',\theta))
  \end{equation}
  which evaluates to 
  \begin{equation}
    \log u_0 = -N(\theta'-\theta)\left(\theta-0.5+\frac{\theta'-\theta}{2}\right) 
  \end{equation}
  and rewrite the test now as
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-\left(\theta+\frac{\theta'-\theta}{2} + \frac{\log u_0}{N(\theta'-\theta)}\right)
    \hspace{0.1in} \not\approx \hspace{0.1in} 0
  \end{equation}
  where $\not\approx$ means ``significantly different from'' under the distribution over samples of $x_i$. 
  The above choice of $u_0$ ensures that the terms in parenthesis above sum to 0.5. Since the $x_i$ have mean 0.5, the
  test will never correctly succeed.

@@

 The choices above for $\theta$ and $\theta'$ guarantee that
  \begin{equation}
    \log u_0 \in -[0.5,1][0.75,1.5] = -[0.375,1.5]
        \end{equation}
and consider the range of $u$ values
  \begin{equation}
    \log u \in \log u_0 + [-0.5,0.375]
  \end{equation}
  the size of the range in $u$ is at least $\exp([-2,-1.125]) = [0082,0.286]$ and occurs
  with probability at least $p_2=0.2044$. With $u$ in this range, we rewrite the test as:
  \begin{equation}
    \frac{1}{b}\sum_{i=1}^b x_i-0.5 \hspace{0.1in} \not\approx \hspace{0.1in} \frac{\log u/u_0}{N(\theta'-\theta)}
  \end{equation}
  so that the LHS has expected value zero.  Given our choice of intervals for the variables, the RHS is in the range $1/\sqrt{N}[-1,0.75]$.
  The standard deviation of the LHS given the interval constraints is at least $0.5/\sqrt{b}$. And
  so the gap between LHS and RHS is at most $2\sqrt{b/N}$ standard deviations.

  The samples $\theta$, $(\theta'-\theta)$
  and $u$ are drawn independently and so the probability of the conjunction of these events is
  $c = p_0 p_1 p_2 = 0.009$.
\end{proof}
There is a separate line of MCMC work drawing principles from
statistical physics. By viewing random variables as particles in a
system, one can apply Hamiltonian Monte Carlo
(HMC)~\cite{mcmc_hamiltonian_2010} methods which generate high
acceptance \emph{and} distant proposals when run on full batches of
data. Recently Langevin
Dynamics~\cite{langevin_2011,conf/icml/AhnBW12} has been applied to
Bayesian estimation on minibatches of data. This simplified dynamics
uses local proposals and avoids MH tests by using small proposal steps
whose acceptance approaches 1 in the limit. However, the constraint on
proposal step size is severe, and the state space exploration reduces
to a random walk. Full minibatch HMC for minibatches was recently
described in ~\cite{sghmc_2014} which allows momentum-augmented
proposals with larger step sizes. However, step sizes are still
limited by the need to run accurately without MH tests.
% \cite{stochastic_thermostats_2014}
Our work opens the door to applying those methods with much
more aggressive step sizes without loss of accuracy. We demonstrate
this in Section~\ref{ssec:nets}.

@@

For our new M-H test, we denote the exact and approximate log likelihood ratios as $\Delta$ and $\Delta^*$.
First, $\Delta$ is defined as:
\begin{equation}\label{eq:delta0}
  \Delta(\theta,\theta') = \log\frac{p(\theta')q(\theta\mid\theta')}{p(\theta)q(\theta'\mid\theta)}
\end{equation}
and in the Bayesian case
\begin{equation}\label{eq:delta1}
\Delta(x,\theta,\theta')  =
\log \frac{p_0(\theta')\prod_{i=1}^N p(x_i \mid \theta')q(\theta \mid
  \theta')}{p_0(\theta)\prod_{i=1}^N p(x_i \mid \theta)q(\theta' \mid\theta)}
  \end{equation}
and we can separate out terms dependent and independent of the data $x$ as:
\begin{equation}\label{eq:delta2}
\Delta(x,\theta,\theta') = N\psi(1,\theta,\theta') + \sum_{i=1}^N\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)}
\end{equation}
A minibatch estimator of $\Delta$ can be defined as:
\begin{equation}\label{eq:delta3}
    \Delta^*(x,\theta,\theta') =
N\psi(1,\theta,\theta') + \frac{N}{b}\sum_{i=1}^b\log\frac{p(x_i\mid\theta')}{p(x_i\mid\theta)}
\end{equation}
Note that $\Delta$ and $\Delta^*$ are evaluated on the full dataset and a minibatch of size $b$
respectively. The scaling term $\frac{N}{b}$ ensures that $\Delta^*(x,\theta,\theta')$ is an
unbiased estimator of $\Delta(x,\theta,\theta')$.

\subsection{A Full Dataset Acceptance Test}\label{ssec:deltas_full}

The key to our test is a smooth acceptance function.  We consider tests other than the Metropolis
test that satisfy the conditions needed for accurate posterior estimation. This condition is
the detailed balance condition defined earlier. Expressed in terms of $\Delta$, the
classical Metropolis test is $f(\Delta) = {\rm min}(\exp(\Delta),1)$. More generally we have:

\begin{lemma}\label{lem:detailed_balance}
  If $g(s)$ is any function such that $g(s) = \exp(s) g(-s)$, then the acceptance function $\alpha(\theta,\theta') \triangleq  g(\Delta(\theta,\theta'))$ satisfies detailed balance. That is,
  \begin{equation}\label{detbal}
    p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta)
    \end{equation}

\end{lemma}

\begin{proof}
  Start with the LHS of (\ref{detbal}), substitute for $\alpha(\theta,\theta')$, and then apply the
  identity $g(s) = \exp(s)g(-s)$:
  \begin{equation}\label{detbal1}
    \begin{split}
      p(\theta)q(\theta'\mid\theta)\alpha(\theta,\theta') =~ & p(\theta)q(\theta'\mid\theta)g(\Delta(\theta,\theta'))\\
      =~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(-\Delta(\theta,\theta'))
    \end{split}
  \end{equation}
  Note then that $-\Delta(\theta,\theta') = \Delta(\theta',\theta)$ because $\Delta$ is the log of ratios of
  terms involving $\theta$ and $\theta'$ and exchanging the two inverts all these ratios, negating the
  logarithm. Continuing:
  \begin{equation}\label{detbal2}
    \begin{split}
    p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(-\Delta(\theta,\theta')) = ~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))g(\Delta(\theta',\theta))\\
    = ~ & p(\theta)q(\theta'\mid\theta)\exp(\Delta(\theta,\theta'))\alpha(\theta',\theta)
    \end{split}
  \end{equation}
  Substituting now for $\Delta(\theta,\theta')$ from equation (\ref{eq:delta0}) gives
  \begin{equation*}\label{detbal3}
    p(\theta)q(\theta'\mid\theta)\frac{p(\theta')q(\theta\mid\theta')}{p(\theta)q(\theta'\mid\theta)}\alpha(\theta',\theta)
      = p(\theta')q(\theta\mid\theta')\alpha(\theta',\theta) = RHS
  \end{equation*}
\end{proof}

As a sanity check, choosing $g(s) = \min\{1, \exp(s) \}$ produces the
standard Metropolis acceptance test $\alpha(\theta,\theta') = g(\Delta(\theta,\theta')) = 
\min\left\{1, \frac{p(\theta')q(\theta \mid
  \theta')}{p(\theta)q(\theta' \mid \theta)}\right\}$. $g(s)$ satisfies
the condition $g(s) = \exp(s)g(-s)$, and therefore detailed
balance. In fact it is the optimal acceptance function in terms of
acceptance rate, since it accepts with probability 1 for $\Delta > 0$.

\subsubsection{Logistic Acceptance Function}\label{ssec:deltas_full}
For our new MH test we use the logistic function: $g(s) =
(1+\exp(-s))^{-1}$. Straightforward arithmetic shows that it satisfies the condition in
Lemma~\ref{lem:detailed_balance}.  While it is slightly less efficient than the Metropolis
test when used on the full dataset, we will see soon that its smoothness allows it to naturally
tolerate substantial variance in its input argument. This in turn will lead to a much more efficient
test on subsets of data.

Assume we begin with the current sample $\theta$  and a
candidate sample $\theta'$. Let $u$ be a uniform random variable $u \sim
\mathcal{U}[0,1]$. We accept $\theta'$ if $g(\Delta(\theta,\theta')) > u$, and reject otherwise.
Since $g(s)$ is monotonically increasing, its inverse $g^{-1}(s)$ is well-defined and
unique. So an equivalent test is to accept $\theta'$ iff
\begin{equation}
  \Delta(\theta,\theta') > v = g^{-1}(u)
\end{equation}
where $v$ is a random variable with the logistic distribution (its CDF is the logistic function). To
see this notice that $\frac{du}{dv} = g'$, that $g'$ is the density corresponding to a logistic
CDF, and finally that $\frac{du}{dv}$ is the density of $v$. The density of $v$ is symmetric,
so we can equivalently test whether
\begin{equation}
  \Delta(\theta,\theta') + v > 0
\end{equation}
for a logistic random $v$.
